---
title: "PAMR main processing"
output:
  html_document:
    self_contained: false
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = here::here("output/"))})
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load data and Preprocessing

Can be examined on data_load_and_preprocess.Rmd/html auxillary notebook, here it is only loaded:

```{r, include=FALSE}
source(here::here("scripts","Rmds","data_load_and_preprocess.R"))
```

# Fitting

```{r, echo = TRUE, results = 'hide', message=FALSE, warning=FALSE}
library(pamr)

pamr=pamr.train(traindata)
adaptive_thresholding=TRUE

  if (adaptive_thresholding) {
    thres=pamr.adaptthresh(pamr)
    pamr=pamr.train(traindata, threshold.scale = thres)
  }

pamrcv=pamr.cv(pamr, traindata)
```

The $\theta$ scaling factors:

```{r}
thres
```

```{r, fig.height=9}
pamrcv
pamr.plotcv(pamrcv)
```


## Threshold λ selection

Selection of two threshold value to evaluate

$To$ - threshold "optimal", lowest cross-validated error rate <br>
$Th$ - threshold "higher" , suboptimal (but still errors do not explode), with fewer genes selected

```{r}
To=17.38  #  Optimal threshold
Th=24.33  #   Higher threshold
```

Visualizing shrunken deltas (differences from shrunken centroid to overall centroids) and saving also the index of the retained genes for classification at the two thresholds (that can be used to run another classification algorithm that does not perform also feature selection):

<details>
<summary>Click to expand</summary>

```{r}
selectedgenes=pamr.listgenes(pamr, traindata, threshold = To)
selectedgenes=as.vector(selectedgenes[,1])
print(paste("Selected genes at optimal threshold to", To, ":", length(selectedgenes)))
sgenes_indexTo <- rownames(traindata$x) %in% selectedgenes
selectedgenes=pamr.listgenes(pamr, traindata, threshold = Th)
selectedgenes=as.vector(selectedgenes[,1])
print(paste("Selected genes at optimal threshold to", To, ":", length(selectedgenes)))
sgenes_indexTh<- rownames(traindata$x) %in% selectedgenes
sgenes_index=data.frame(sgenes_indexTo,sgenes_indexTh)
write.csv(sgenes_index, here::here("output/selected_genes.csv"), row.names = FALSE)
```

</details>


# Visualizing classifications

Loading required libraries:

```{r, echo=TRUE, include=FALSE}
library(classmap) 
library(classmapExt) #The novel auxillary package to classmap
```

### Training Set

Using vcr.pamr.train function to produce all the quantities needed for the visualization of the two models with the two thresholds:

```{r}
vcrtrainTo=vcr.pamr.train(data=traindata, pamrfit = pamr, threshold=To)
vcrtrainTh=vcr.pamr.train(data=traindata, pamrfit = pamr, threshold=Th)
```

#### Silhouttes

```{r, dev='svg'}
cfm=caret::confusionMatrix(factor(vcrtrainTo$pred),factor(vcrtrainTo$y))
cfm$table
silplot(vcrtrainTo, main=paste0("(Training) Silplot at threshold: ",To))
```

```{r, dev='svg'}
cfm=caret::confusionMatrix(factor(vcrtrainTh$pred),factor(vcrtrainTh$y))
cfm$table
silplot(vcrtrainTh, main=paste0("(Training) Silplot at threshold: ",Th))
```

#### Classmaps

```{r , fig.height=3.5, dev='svg'}
par(mfrow=c(1,3))
classmap(vcrtrainTo, whichclass = 1, main = "Pred of class LUAD" )
classmap(vcrtrainTo, whichclass = 2, main = "Pred of class LUSC")
classmap(vcrtrainTo, whichclass = 3, main = "Pred of class NORM")
mtext(paste0("Classmaps for training at threshold: ",To), line=-5, side=3, outer=TRUE, cex=1)
```

```{r , fig.height=3.5, dev='svg'}
par(mfrow=c(1,3))
classmap(vcrtrainTh, whichclass = 1, main = "Pred of class LUAD" )
classmap(vcrtrainTh, whichclass = 2, main = "Pred of class LUSC")
classmap(vcrtrainTh, whichclass = 3, main = "Pred of class NORM")
mtext(paste0("Classmaps for training at threshold: ",Th), line=-5, side=3, outer=TRUE, cex=1)
```

#### MDS color-scaled plots

```{r}
mdscolorscale(vcrtrainTo, diss=vcrtrainTo$pwd, main=
                paste0("(Train) MDScolorscale at thresh: ",To))
```

```{r}
mdscolorscale(vcrtrainTh, diss=vcrtrainTh$pwd, main=
                paste0("(Train) MDScolorscale at thresh: ",Th))
```

#### Quasi Residual Plots

Drawn only for higher threshold $Th$

For some continuos covariates:

```{r, fig.height=8, fig.width=10, dev='svg'}

par(mfrow=c(2,3))

qresplot(vcrtrainTh$PAC,traindata$pancovariates$MSI.MANTIS.Score, plotErrorBars = TRUE, 
         main="Mantis Score", xlim=c(0.2,0.4))
qresplot(vcrtrainTh$PAC,traindata$covariates$years_smoked, plotErrorBars = TRUE, 
         main="Years smoked")
qresplot(vcrtrainTh$PAC,traindata$covariates$longest_dimension, plotErrorBars = TRUE, 
         main="Tumor dimension (longest)", xlim=c(0,3))
qresplot(vcrtrainTh$PAC,traindata$pancovariates$Fraction.Genome.Altered, plotErrorBars = TRUE, 
         main="Fraction of Genome Altered")
qresplot(vcrtrainTh$PAC,traindata$pancovariates$Diagnosis.Age, plotErrorBars = TRUE, 
         main="Age")
qresplot(vcrtrainTh$PAC,traindata$covariates$`Alignment Quality Index`, plotErrorBars = TRUE,
         main="Alignment Quality Index", xlim = c(0.045,0.11), grid=seq(0.04, 0.11, length.out=10))

```

For some discrete covariates oppurtunely integer-coded ( a table with corresponding values is provided)

```{r, fig.height=8, dev='svg'}

legend=data.frame(matrix(nrow = 20, ncol = 0))
legend$Code=1:20

par(mfrow=c(2,2))

#ajcc stage
discrete_covariate=factor(unlist(traindata$covariates$ajcc_pathologic_stage, use.names = FALSE)) 
levels=levels(discrete_covariate)
discrete_cov_length=length(levels)
if (length(levels) < 20) {
    levels <- c(levels, rep(NA, 20 - length(levels)))
}
legend$`Ajcc Stage`=levels
qresplot(vcrtrainTh$PAC,as.numeric(discrete_covariate), plotErrorBars = TRUE,
         grid=seq(1:(1+discrete_cov_length))-0.5 , main = "Ajcc Stage")
#histologic type
discrete_covariate=factor(unlist(traindata$covariates$morphology, use.names = FALSE)) 
levels=levels(discrete_covariate)
discrete_cov_length=length(levels)
if (length(levels) < 20) {
    levels <- c(levels, rep(NA, 20 - length(levels)))
}
legend$`Histologic Type`=levels
qresplot(vcrtrainTh$PAC,as.numeric(discrete_covariate), plotErrorBars = TRUE,
         grid=seq(1:(1+discrete_cov_length))-0.5 , main = "Histologic Type")
#site
discrete_covariate=factor(unlist(traindata$covariates$site_of_resection_or_biopsy, use.names = FALSE)) 
levels=levels(discrete_covariate)
discrete_cov_length=length(levels)
if (length(levels) < 20) {
    levels <- c(levels, rep(NA, 20 - length(levels)))
}
legend$`Site`=levels
qresplot(vcrtrainTh$PAC,as.numeric(discrete_covariate), plotErrorBars = TRUE,
         grid=seq(1:(1+discrete_cov_length))-0.5 , main = "Site")
#gender
discrete_covariate=factor(unlist(traindata$covariates$gender, use.names = FALSE)) 
levels=levels(discrete_covariate)
discrete_cov_length=length(levels)
if (length(levels) < 20) {
    levels <- c(levels, rep(NA, 20 - length(levels)))
}
legend$`Gender`=levels
qresplot(vcrtrainTh$PAC,as.numeric(discrete_covariate), plotErrorBars = TRUE,
         grid=seq(1:(1+discrete_cov_length))-0.5 , main = "Gender", xlim=c(0,3))


library(kableExtra)
knitr::kable(legend, "html", align = "c") %>% 
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = F)
```

On selected genes for the classification in dataset:

```{r, fig.height=10, fig.widht=10, dev='svg'}

selectedgenes=which(sgenes_index$sgenes_indexTh)

#in all the genes
coeff=rep(0,nrow(traindata$x[selectedgenes,]))

#getting indexes only of genes that take on more that 7 unique values (<7 values it is rather discrete)
#also lefting out genes that are more than 35% observation that take 0 values
index_nonsparse <- which(apply(traindata$x[selectedgenes,], 1, function(x) {
    unique_count <- length(unique(x))
    zero_count <- sum(x == 0)
    return(unique_count >= 7 & zero_count / length(x) < 0.35)}))

for (i in index_nonsparse) {
  lm=lm(vcrtrainTh$PAC~traindata$x[i,])
  coeff[i]=lm$coefficients[2]
}

sorted_indices <- order(coeff, decreasing = TRUE)
sorted_coeff <- sort(coeff, decreasing = TRUE)
sorted_indices[1:9]
sorted_coeff[1:9]

par(mfrow=c(3,3))

for (i in sorted_indices[1:9]) {
  gene=traindata$x[i,]
  qresplot(vcrtrainTh$PAC,gene,plotErrorBars = TRUE, main = rownames(traindata$x)[i])
}
```

On all genes in dataset:

```{r, fig.height=10, fig.width=10, dev='svg'}
#in all the genes
coeff=rep(0,nrow(traindata$x))

#getting indexes only of genes that take on more that 7 unique values (<7 values it is rather discrete)
index_nonsparse <- which(apply(traindata$x, 1, function(x) length(unique(x)) >= 8)) 

#getting indexes only of genes that take on more that 7 unique values (<7 values it is rather discrete)
#also lefting out genes that are more than 35% observation that take 0 values
index_nonsparse <- which(apply(traindata$x, 1, function(x) {
    unique_count <- length(unique(x))
    zero_count <- sum(x == 0)
    return(unique_count >= 7 & zero_count / length(x) < 0.35)}))

for (i in index_nonsparse) {
  lm=lm(vcrtrainTh$PAC~traindata$x[i,])
  coeff[i]=lm$coefficients[2]
}

sorted_indices <- order(coeff, decreasing = TRUE)
sorted_coeff <- sort(coeff, decreasing = TRUE)
sorted_indices[1:9]
sorted_coeff[1:9]


par(mfrow=c(3,3))

for (i in sorted_indices[1:9]) {
  gene=traindata$x[i,]
  qresplot(vcrtrainTh$PAC,gene,plotErrorBars = TRUE, main = rownames(traindata$x)[i], 
           grid=seq(0.0, 0.3, length.out=10), xlim = c(0,0.25))
}
```

